---
title: "Homework 3"
author: "Lauren Brochu (leb3846) - SDS315 UT Austin"
date: "27 January 2025"
output:
  pdf_document: default
  html_document: default
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.height=3, fig.width=4, warning=FALSE, tidy=TRUE, echo=FALSE, tidy.opts=list(width.cutoff=60))
```

```{r}
# read in datasets 
gasPrices <- read.csv('gasprices.csv')
sclass <- read.csv('sclass.csv')
pilotSurvey <- read.csv('nbc_pilotsurvey.csv')
ebay <- read.csv('ebay.csv')
```

```{r, message = FALSE}
# load in libraries
library(ggplot2)
library(dplyr)
library(boot)
library(mosaic)
library(knitr)

```

## [GitHub Link !!](https://github.com/laurenbrochu/2025SDS315/blob/main/SDS315.Homework3.Rmd)
\newpage

# Problem 1 

## Theory A --- Gas Stations Without Competitors Charge More

```{r, fig.width = 8, fig.height = 4}
# bootstrap by price and competitors 
gasPartaBootComp = do(10000)*diffmean(Price ~ Competitors, data = resample(gasPrices))

# display 
ggplot(gasPartaBootComp) + 
  geom_histogram(aes(x=diffmean), bins = 30, 
                 col = 'navy', fill = 'lightblue') + 
  labs(
       x = 'Difference in Mean Price',
       y = 'Frequency') +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"), 
        axis.title.x = element_text(size = 12, face = "bold"), 
        axis.title.y = element_text(size = 12, face = "bold"))

# CI
confint(gasPartaBootComp, level = 0.95) %>%
  kable(caption = "95% Confidence Interval Statistics")


```

Claim: Gas stations charge more if they lack direct competition in sight. 
\newline
\newline
Evidence: The difference in gas prices with and without competition are shown by their bootstrapped mean data. This discrepancy is approximately -0.023 across a 95% confidence interval of about -0.0557704 and 0.0073182. 
\newline
\newline
Conclusion: The data does not appear to support the claim that gas stations without direct competition charge more because the confidence interval includes zero. There's likely no statistically significant evidence that gas stations in this sample support the claim. 

\newpage

## Theory B --- The Richer the Area, the Higher the Gas Prices
```{r, fig.width = 8, fig.height = 4}
# bootstrap by price and income 
gasPartbBootIncome = do(10000) * cor(Price ~ Income, data = resample(gasPrices))

#display
ggplot(gasPartbBootIncome) + 
  geom_histogram(aes(x=cor), bins = 30, 
                 col = 'navy', fill = 'lightblue') + 
  labs(
       x = 'Difference in Mean Price',
       y = 'Frequency') +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"), 
        axis.title.x = element_text(size = 12, face = "bold"), 
        axis.title.y = element_text(size = 12, face = "bold"))

# CI
confint(gasPartbBootIncome, level = 0.95) %>%
  kable(caption = "95% Confidence Interval Statistics")

```

Claim: The richer the area, the higher the gas prices. 
\newline
\newline
Evidence: The difference in gas prices across wealthier areas is shown by their bootstrapped mean data. This discrepancy is approximately 0.396 across a 95% confidence interval of about 0.2035 and 0.5681, indicating a moderate positive correlation. 
\newline
\newline
Conclusion: There appears to be a statistically significant correlation between wealthier areas and gas prices, as the data distribution does not include 0. The data likely supports the claim, meaning gas stations in wealthier areas may set higher prices. 

\newpage

## Theory C --- Gas Stations at stoplights charge more

```{r, fig.width = 8, fig.height = 4}
# bootstrap by price and stoplight
gasPartcBootStoplight = round(do(10000)*diffmean(Price ~ Stoplight, data = resample(gasPrices)), 3)

#display
ggplot(gasPartcBootStoplight) + 
  geom_histogram(aes(x=diffmean), bins = 30, 
                 col = 'navy', fill = 'lightblue') + 
  labs(
       x = 'Difference in Mean Price',
       y = 'Frequency') +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"), 
        axis.title.x = element_text(size = 12, face = "bold"), 
        axis.title.y = element_text(size = 12, face = "bold"))

#CI
confint(gasPartcBootStoplight, level = 0.95) %>%
  kable(caption = "95% Confidence Interval Statistics")

```

Claim: Gas stations at stoplights charge more. 
\newline
\newline
Evidence: The difference in gas prices in relation to stoplights are shown by their bootstrapped mean data. This discrepancy is approximately -0.0033 across a 95% confidence interval of about -0.038 and 0.031. 
\newline
\newline
Conclusion: The data does not appear to support the claim that gas stations charge more in proximity to stoplights because the confidence interval includes zero. There’s likely no statistically significant evidence supporting this relationship.

\newpage

## Theory D --- Gas Stations with Direct Highway Access Charge More

```{r, fig.width = 8, fig.height = 4}
# bootstrap by price and highway
gasPartdBootHighway = do(10000)*diffmean(Price ~ Highway, data = resample(gasPrices))

#display
ggplot(gasPartdBootHighway) + 
  geom_histogram(aes(x=diffmean), bins = 30, 
                 col = 'navy', fill = 'lightblue') + 
  labs(
       x = 'Difference in Mean Price',
       y = 'Frequency') +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"), 
        axis.title.x = element_text(size = 12, face = "bold"), 
        axis.title.y = element_text(size = 12, face = "bold"))

#CI
confint(gasPartdBootHighway, level = 0.95) %>%
  kable(caption = "95% Confidence Interval Statistics")

```

Claim: Gas stations with direct highway access charge more. 
\newline
\newline
Evidence: The difference in gas prices in relation to highway access are shown by their bootstrapped mean data. This discrepancy is approximately 0.0457 across a 95% confidence interval of about 0.0087 and 0.081. 
\newline
\newline
Conclusion: There appears to be a statistically significant correlation between highway access and gas prices, as the data distribution does not include 0. This supports the claim that gas stations with highway access tend to charge more.

\newpage

## Theory E --- Shell Charges More Than All Other Non-Shell Brands

```{r, fig.width = 8, fig.height = 4}
# bootstrap by shell and non shell stations 
gasPrices$gasShell = ifelse(gasPrices$Brand == 'Shell', 'Shell', 'Other')
gasParteBootShell = do(10000) * diffmean(Price ~ gasShell, data = resample(gasPrices))

# display
ggplot(gasParteBootShell) + 
  geom_histogram(aes(x=diffmean), bins=30, 
                 col = 'navy', fill = 'lightblue') + 
  labs(
       x = 'Difference in Mean Price',
       y = 'Frequency') +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"), 
        axis.title.x = element_text(size = 12, face = "bold"), 
        axis.title.y = element_text(size = 12, face = "bold"))

# CI
confint(gasParteBootShell, level = 0.95) %>%
  kable(caption = "95% Confidence Interval Statistics")

```

Claim: Shell charges more than all other non-Shell brands. 
\newline
\newline
Evidence: The difference in gas prices across Shell and non-Shell providers is shown by their bootstrapped mean data. This discrepancy is approximately 0.0274 across a 95% confidence interval of about -0.0102 and 0.0641.
\newline
\newline
Conclusion: The data does not appear to support the claim that Shell gas stations charge more than non-Shell companies because the confidence interval includes zero. There's likely no statistically significant evidence that gas stations in this sample support the claim.

\newpage

# Problem 2
## Part A: Average Milage of 2011 S-Class 63 AMG

```{r, fig.width = 8, fig.height = 4}
# filter data by year and trim
sclassAMG = sclass %>% 
  filter(year == 2011, trim == '63 AMG')
sclassPartaBootAMG <- do(10000) * mean(~mileage, data = resample(sclassAMG))

# display
ggplot(sclassPartaBootAMG) +
  geom_histogram(aes(x= mean), bins=30, 
                 col = 'navy', fill = 'lightblue') + 
  labs(
       x = 'Difference in Mileage',
       y = 'Frequency') +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"), 
        axis.title.x = element_text(size = 12, face = "bold"), 
        axis.title.y = element_text(size = 12, face = "bold"))

# CI
confint(sclassPartaBootAMG, level = 0.95) %>% 
  kable(caption = "95% Confidence Interval Statistics")
```

\newpage

## Part B: Proportion of Black 2014 S-Class 550s

```{r, fig.width = 8, fig.height = 4}
# fresh data
sclassPartbBlack <- read.csv('sclass.csv')

# filter and mutate for color and model
sclassPartbBlack = sclassPartbBlack %>% 
  filter(year == 2014, trim == '550') %>%
  mutate(isBlack = ifelse(color == 'Black', yes = 'Black', no = 'Not Black'))

# bootstrap 
sclassPartbBootBlack <- do(10000) * prop(~isBlack, data =  resample(sclassPartbBlack))

# display
ggplot(sclassPartbBootBlack) +
  geom_histogram(aes(x= prop_Black), bins=30, 
                 col = 'navy', fill = 'lightblue') + 
  labs(
       x = 'Difference in Mileage',
       y = 'Frequency') +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"), 
        axis.title.x = element_text(size = 12, face = "bold"), 
        axis.title.y = element_text(size = 12, face = "bold"))

# CI
confint(sclassPartbBootBlack, level = 0.95) %>% 
  kable(caption = "95% Confidence Interval Statistics")

```
\newpage

# Problem 3

## Part A --- *Living with Ed* and *My Name is Earl* on Happiness
```{r, fig.width = 8, fig.height = 4}
# fresh data
nbcHappiness  = read.csv('nbc_pilotsurvey.csv')
# filter by show
nbcHappiness = nbcHappiness %>% 
  filter(Show == 'My Name is Earl' | Show == 'Living with Ed')

# bootstrap
nbcPartaBoot <- do(10000) * diffmean(Q1_Happy ~ Show, data = resample(nbcHappiness))

# display
ggplot(nbcPartaBoot) +
    geom_histogram(aes(x= diffmean), bins=30, 
                   col = 'navy', fill = 'lightblue') + 
  labs(
       x = 'Difference in Response',
       y = 'Frequency') +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"), 
        axis.title.x = element_text(size = 12, face = "bold"), 
        axis.title.y = element_text(size = 12, face = "bold"))

# CI
confint(nbcPartaBoot, level = 0.95) %>% 
  kable(caption = "95% Confidence Interval Statistics")
```

Question: Is there evidence that one show consistently makes viewers happier?
\newline
\newline
Approach: After filtering, I bootstrapped the mean differences between happy responses between *Living with Ed* and *My Name is Earl*. 
\newline
\newline
Results: The mean difference in happy responses between the two shows was about -0.1491, with a 95% confidence interval of -0.3991 and 0.104.
\newline
\newline
Conclusion: There doesn't appear to be statistically significant evidence that one show consistently makes viewers more happy given that zero is included in the distribution. The claim is not supported in this sample. 
\newline
\newline

\newpage

## Part B --- *The Biggest Loser* and *The Apprentice: Los Angeles* on Annoyance 
```{r, fig.width = 8, fig.height = 4}
# fresh data
nbcAnnoyance  = read.csv('nbc_pilotsurvey.csv')

# filter by show
nbcAnnoyance = nbcAnnoyance %>% 
  filter(Show == 'The Biggest Loser' | Show == 'The Apprentice: Los Angeles')

# bootstrap
nbcPartbBoot <- do(10000) * diffmean(Q1_Annoyed ~ Show, data = resample(nbcAnnoyance))

# display
ggplot(nbcPartbBoot) +
    geom_histogram(aes(x= diffmean), bins=30,
                   col = 'navy', fill = 'lightblue') + 
  labs(
       x = 'Difference in Response',
       y = 'Frequency') +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"), 
        axis.title.x = element_text(size = 12, face = "bold"), 
        axis.title.y = element_text(size = 12, face = "bold"))
# CI
confint(nbcPartbBoot, level = 0.95)  %>% 
  kable(caption = "95% Confidence Interval Statistics")
```
Question: Is there evidence that one show consistently makes viewers more annoyed?
\newline
\newline
Approach: After filtering, the mean differences between annoyed responses between *The Biggest Loser* and *The Apprentice: Los Angeles* were gathered by bootstrapping. 
\newline
\newline
Results: A difference of about -0.271 was produced from the bootstrap of the two shows within a 95% confidence interval of -0.5258 and -0.016. 
\newline
\newline
Conclusion: Because zero is not in the interval, there is likely no statistically significant indicators that reviews found *The Apprentice: Los Angeles* more annoying to viewers.  
\newline
\newline

\newpage 

## Part C --- *Dancing with the Stars* on Confusion
```{r, fig.width = 8, fig.height = 4}
# fresh data
nbcConfusion  = read.csv('nbc_pilotsurvey.csv')

# filter for DWTS
nbcConfusion = nbcConfusion %>% 
  filter(Show == 'Dancing with the Stars')
nbcPartcBoot <- do(10000) * prop(~Q2_Confusing > 3, data = resample(nbcConfusion))

# display
ggplot(nbcPartcBoot) +
    geom_histogram(aes(x = prop_TRUE), bins=30, 
                   col = 'navy', fill = 'lightblue') + 
  labs(
       x = 'Difference in Response',
       y = 'Frequency') +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"), 
        axis.title.x = element_text(size = 12, face = "bold"), 
        axis.title.y = element_text(size = 12, face = "bold"))

# CI
confint(nbcPartcBoot, level = 0.95) %>% 
  kable(caption = "95% Confidence Interval Statistics")

```
Question: What proportion of of American TV watchers would we expect to give a response of 4 or greater to indicate that "Dancing With the Stars" is a confusing show?
\newline
\newline
Approach: A bootstrapped proportion of confused responses was gathered after filtering the show's data. 
\newline
\newline
Results: A difference of about 0.0773 was produced from the bootstrapped proportion of the show within a 95% confidence interval of 0.0387 and 0.1160. 
\newline
\newline
Conclusion: The proportion of Americans who would likely rate "Dancing with the Stars" as confusing via a rating of four or more is between 0.0387 and 0.1160 with 95% confidence.  
\newline
\newline

\newpage

# Problem 4 --- Ebay

```{r, fig.width = 8, fig.height = 4}
# find difference in revenue before and after Google ads 
ebay = ebay %>% 
  mutate(ratio = (rev_after / rev_before))

# bootstrap
ebayBoot = do(10000) * diffmean(ratio ~ adwords_pause, data = resample(ebay))

# display
ggplot(ebayBoot) +
    geom_histogram(aes(x = diffmean), bins=30, 
                   col = 'navy', fill = 'lightblue') + 
  labs(
       x = 'Difference in Response',
       y = 'Frequency') +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"), 
        axis.title.x = element_text(size = 12, face = "bold"))

# CI
confint(ebayBoot, level = 0.95) %>% 
  kable(caption = "95% Confidence Interval Statistics")

```

Question: Does the extra traffic brought to Etsy from paid search results justify the cost of the ads themselves?
\newline
\newline
Approach: I determined potential increase in revenue by finding the ratio of pre- and post-paid-advertisements. I then bootstrapped by DMA indicator.
\newline
\newline
Results: The mean difference in revenue before and after the pause was about -0.05228 within 95% certainty between -0.0909 and -0.0133. 
\newline
\newline
Conclusion: Statistically significant evidence exists that pausing paid advertisements was correlated with a smaller revenue ratio, supported by the lack of zero in the distribution. This suggests that paid search results had a positive impact on Etsy’s revenue. 
