---
title: "SDS315 HW 8"
author: "Lauren Brochu (EID - leb3846)"
date: "14 April 2025"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.height=8, fig.width= 12, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=60), message=FALSE)
```

# [GitHub Link](https://github.com/laurenbrochu/2025SDS315/blob/main/SDSHW7.Rmd)

\newpage

## Problem 1: Regression Warm Up 

```{r, echo = FALSE}

# read in data 
creatine <- read.csv("creatinine.csv")

```

```{r, echo = FALSE}

# linear regression 
creatineLR <- lm(creatclear ~ age, data = creatine)

```

```{r, include = FALSE}

# a) clearance for 55 y/o
predict(creatineLR, newdata = data.frame(age = 55))

```

The estimated clearance rate for a 55 year old is predicted to be 113.723 according to the linear regression model constructed from the _creatclear_ and _age_ variables. This means that we would expect a creatine clearance rate of 113.72 mL/min for a 55 year old. 

```{r, include = FALSE}

# b) clearance rate with age 
coef(creatineLR)

```
 
An r value of -0.62 suggests that as age increases year by year, the creatine clearance rate decreased by approximately 0.62 mL/min. The negative relationship of the slope coefficient confirms that kidney function generally declines with age.

```{r, include = FALSE}

# c) 40 vs 60 y/0 
creatineIntercept <- 147.8129
creatineSlope <- -0.6198159

# exp value
fortyExp <- creatineIntercept + creatineSlope * 40
sixtyExp <- creatineIntercept + creatineSlope * 60

# obs value 
fortyObs <- 135 - fortyExp
sixtyObs <- 112 - sixtyExp

```
 
When comparing health of a 40 year old with a clearance of 135 and a 60 year old with a clearance of 112, it was found that while both people were of higher clearance than expected, the former's residual was much higher, suggesting better health for their age: 
 
* **40 year old**: 
  + Expected Clearance: 123.02
  + Actual Clearance: 135
  + Exceeded Model by 11.98
 
* **60 year old**: 
  + Expected Clearance: 110.62
  + Actual Clearance: 112
  + Exceeded Model by 1.38

\newpage

## Problem 2: Modeling Disease Growth

```{r, include = FALSE}

# read in data
covid <- read.csv("covid.csv")

# load libraries 
library(ggplot2)
library(dplyr)
library(mosaic)

```

```{r, include = FALSE}

# Filter data for each country
italy <- covid %>% filter(country == "Italy")
spain <- covid %>% filter(country == "Spain")

# Create log of deaths
italy$logDeath <- log(italy$deaths)
spain$logDeath <- log(spain$deaths)

# Fit linear models on log(deaths)
italyModel <- lm(logDeath ~ days_since_first_death, data = italy)
spainModel <- lm(logDeath ~ days_since_first_death, data = spain)

# Estimated growth rates
italyGrowth <- coef(italyModel)[2]
spainGrowth <- coef(spainModel)[2]

# Estimated doubling times
italyDT <- log(2) / italyGrowth
spainDT <- log(2) / spainGrowth

# Bootstrap for Italy
bootItaly <- do(10000) * {
  sample <- mosaic::resample(italy)
  coef(lm(logDeath ~ days_since_first_death, data = sample))[2]
}
italyCI <- confint(bootItaly, level = 0.95)

# Bootstrap for Spain
bootSpain <- do(10000) * {
  sample <- mosaic::resample(spain)
  coef(lm(logDeath ~ days_since_first_death, data = sample))[2]
}
spainCI <- confint(bootSpain, level = 0.95)

# Convert growth rate CIs to doubling time CIs
italyCI_vals <- as.numeric(italyCI)
spainCI_vals <- as.numeric(spainCI)

italyDT_CI <- log(2) / rev(italyCI_vals)
spainDT_CI <- log(2) / rev(spainCI_vals)


# Round as specified
italyGrowth <- round(italyGrowth, 3)
spainGrowth <- round(spainGrowth, 3)
italyDT <- round(italyDT, 1)
spainDT <- round(spainDT, 1)
italyCI_vals <- round(as.numeric(italyCI[1, c("lower", "upper")]), 3)
spainCI_vals <- round(as.numeric(spainCI[1, c("lower", "upper")]), 3)
italyCI_vals <- as.numeric(unlist(italyCI[ , c("lower", "upper")]))
spainDT_CI <- round(log(2) / rev(spainCI_vals), 1)

```

```{r, echo = FALSE}
# Plot the daily deaths
ggplot(covid, aes(x = days_since_first_death, y = deaths, color = country)) +
  geom_line(size = 1) +
  labs(
    title = "Daily COVID-19 Deaths in Italy and Spain",
    x = "Days Since First Death",
    y = "Reported Deaths"
  ) +
  theme_minimal()  + 
  theme(
    plot.title = element_text(face = "bold", size = 18),
    axis.title = element_text(face = "bold", size = 14), 
    legend.title = element_text(face = "bold", size = 14))


```

A linear regression model was made to that modeling the early exponential growth of COVID-19 deaths was possible post-log transformation of the _days_since_first_death_ variable. This also allowed is to estimate growth rates and doubling times.  

* **Italy** 
  + Growth Rate: 0.183
    + 95% Confidence Interval for Growth Rate: (0.159, 0.208)
  + Doubling Time: 3.8 Days
    + 95% Confidence Interval for Doubling Time: (3.3, 4.4)
 
* **Spain**
  + Growth Rate: 0.276
    + 95% Confidence Interval for Growth Rate: (0.236, 0.318)
  + Doubling Time: 2.5 days
    + 95% Confidence Interval for Doubling Time: (2.2, 2.9)
    
These results prove that during the initial COVID wave, the virus spread more quickly in Spain than in Italy, as supported by the graph of reported deaths over time since Spain's line is steeper.

\newpage

## Problem 3: Price Elasticity of Demand

```{r, include = FALSE} 

# read in data 
milk <- read.csv("milk.csv")

```

```{r, include = FALSE}

# log transform
milk$logPrice <- log(milk$price)
milk$logSales <- log(milk$sales)

# linear model
milkElasticity <- lm(logSales ~ logPrice, data = milk)
summary(milkElasticity)

# estimated elasticity 
estElasticity <- coef(milkElasticity)[2]

# bootstrap
set.seed(42)
B <- 1000
bootBetas <- numeric(B)

for(i in 1:B) { 
  sampleRows <- sample(nrow(milk), replace = TRUE)
  bootModel <- lm(logSales ~ logPrice, data = milk[sampleRows, ])
  bootBetas[i] <- coef(bootModel)[2]}

# confidence interval 
milkCI <- quantile(bootBetas, c(0.025, 0.975))

```

A linear regression was needed to estimate price elasticity directly, so prior to analysis, a log transformation was applied to the _price_ and _sales_ variables. Doing so, the approximate price elasticity was about -1.62, meaning that a 1% increase in price leads to a 1.62% decrease in quantity demand. The confidence interval for elasticty was (-1.78, -1.46), proving that the demand for milk is elastic (consumers are usually responsive to price changes). 



